### 2018-01-09 ident/docs/ident_manuscript.tex
additions to refs and intro section

\begin{document}
\paragraph{Outline:}
\begin{itemize}
	\item Importance of parameter values for estimating in vivo response using kinetic models of metabolism and consequently for metabolic network design using kinetic models of metabolism
	\item The need for parameter identifiability to determine unique and true parameter values from observed data
	\item Types and purpose of identifiability for parameters
	\item Methods for structural identifiability and existing methods for practical identifiability
	\item Lack of methods for practical identifiability and consequently experimental design (not covered?)
	\item work done in this paper for practical identifiability			
	\item scalability of computer algebra-based methods for structural identifiability (using CRNT to reduce networks to make structural identifiability scalable) (move to discussion - may be in discussion)
\end{itemize}
\section{Introduction:}
Kinetic models of metabolism can be used to study the dynamic characteristics of metabolic networks. In these models, ordinary differential equations (ode) are used to express the rate of change of metabolite concentrations ($x$) as a function of the reaction fluxes ($v$) in the metabolic network (Equation \ref{eq:kinstoich}). The matrix $\mathbf{S}$ in Equation (\ref{eq:kinstoich}a) defines the stoichiometric relationship between the fluxes and the concentrations of the metabolic network.
\begin{subequations}\label{eq:kinstoich}
	\begin{align}
	\dot{x} = \mathbf{S}v\\
	v = g(x, \theta, u)
	\end{align}
\end{subequations}
The expression for the nonlinear function ($g$) used to describe each reaction flux $v_i$ in a kinetic model (Equation \ref{eq:kinstoich}b) is dependent on the enzyme kinetic mechanism that is used to model the reaction (older kinetic modeling papers, reviews, Srinivasan et al., 2015). Accordingly, $g$ is a nonlinear function of the metabolite concentrations, enzyme kinetic parameters ($\theta$) and other input concentrations ($u$).

The ability to predict the steady state and dynamic responses of metabolic networks, under in vivo conditions, to different perturbations is dependent on the numerical values of the enzyme kinetic parameter values ($\theta$) in Equation (\ref{eq:kinstoich}). Analyzing the ability of a metabolic network to exhibit dynamic characteristics like multiple steady states and oscillations, irrespective of the structure of the network, is one example where parameter values might play a crucial role (Srinivasan et al., 2017?). The use of in vitro, or unreliable in vivo parameter estimates, reduces confidence in the model predicted behaviour. Consequently, the reduction in confidence hampers the use of these models to gain insight into the functioning of metabolic networks (ensemble modeling and ORACLE papers). The insights gained from the use of kinetic models are subsequently used to design changes to these metabolic networks to achieve various goals. These goals could either be to increase metabolite production for biosynthesis of different chemicals (kinetic model for BDO in ecoli, k-ecoli457, Srinivasan et al., 2017) or to find therapeutic targets that can cut-off metabolic pathways to kill cancer cells (new paper on using geneMCS for cancer cell lines). Hence, an increase in uncertainty in model predicted responses is also an obstacle for using the predicted responses as a basis for designing the metabolic networks to achieve these goals.

If all intracellular metabolite concentrations can be measured over a time course, and the corresponding data is available, a nonlinear programming problem can be formulated to estimate the enzyme kinetic parameters ($\theta$) in Equation (\ref{eq:kinstoich}). The minimization of least square error between the measured ($x^*$) and modeled ($x$) concentrations can be used as an objective function (Equation \ref{eq:lsqopt}a) for the optimization problem (Equation \ref{eq:lsqopt}), with fixed upper ($\theta_u$) and lower ($\theta_l$) bounds for the parameter values (Equation \ref{eq:lsqopt}b). The difference between the data and the model estimate for each output and at each time point in the objective function is weighted by the variance in the experimental data $\sigma_{kl}^*$ for that corresponding variable and time point.
\begin{subequations}\label{eq:lsqopt}
	\begin{align}
	\underset{\theta}{\mathrm{min}} &\sum_{k=1}^{m}\sum_{l=1}^{d}\left(\frac{x_{kl}^*-x_{kl}}{\sigma_{kl}^*}\right)^2\\
	&\theta_l \le \theta \le \theta_u
	\end{align}
\end{subequations}

However, not all metabolite concentrations used in the model (Equation \ref{eq:kinstoich}) can be measured. Additionally, measurable fluxes in the metabolic network also need to be included as part of the parameter estimation problem. In such scenarios, the parameter estimation problem is modified to suit a new system of equations shown below (Equation \ref{eq:output}). The new system of equations is obtained by augmenting the original system (Equation \ref{eq:kinstoich}) with Equation (\ref{eq:output}c) that models the relationship between the measurable metabolite concentrations and fluxes, represented by the variable $y$, and the unmeasured concentrations ($x$) that are used in the original model (Equation \ref{eq:kinstoich}) above. Equation (\ref{eq:output}) now has additional parameters $\mu$ that define this relationship and also need to be estimated.
\begin{subequations}\label{eq:output}
	\begin{align}
	\dot{x} = \mathbf{S}v\\
	v = g(x, y, \theta, u)\\
	\dot{y} = h(x, y, \mu, u)
	\end{align}
\end{subequations}
In systems identification, the measured concentrations and fluxes, $y$ in Equation (\ref{eq:output}c), are called output or observed variables, and the unmeasured concentrations, $x$ in Equation (\ref{eq:output}a-b), are called the state variables. 
For estimating both $\theta$ and $\mu$, the metabolite concentrations ($x$) in the optimization problem (Equation \ref{eq:lsqopt}) are substituted with the output variables $y$.

However, the ability to determine unique solutions to parameters $\theta$ and $\mu$ is governed by the identifiability of these parameters in the model \parencite{McLean2012}. The identifiability of parameters in nonlinear models can be classified into two categories: structural (or a priori) and practical (or posterior) identifiability. 	
Any system (Equation \ref{eq:output}) is said to be structurally identifiable if, for an input-output mapping defined by $\mathbf{y} = \Phi(\mu,u)$ for at least one input function $u$, any two values of parameters $\mathbf{\mu}_1$ and $\mathbf{\mu}_2$ satisfy the relationship in Equation (\ref{eq:stident}) below.
\begin{align}\label{eq:stident}
\Phi(\mathbf{\mu}_1,u) = \Phi(\mathbf{\mu}_2,u) \iff \mathbf{\mu}_1 = \mathbf{\mu}_2
\end{align}
Accordingly, any system that has an infinite number of solutions to the parameter estimation problem for all input functions is said to structurally non-identifiable. Thus, the structural identifiability of parameters in a dynamic model helps establish the presence or absence of a relationship between the unobservable state variables and the observable output variables. Accordingly, the effect of model structure and parameterization on the ability to infer true parameter values from experimental data is determined by the structural identifiability of the parameter.

Experimental data from many physical systems is usually noisy, and when parameters are estimated on the basis of noisy data, the ability to estimate unique parameter values to satisfy Equation (\ref{eq:stident}) is referred to as practical identifiability. The effect of the available experimental data on the ability to estimate unique parameter values is determined by the practical identifiability of the parameter. Accordingly, practical identifiability of a parameter is contingent upon the nature, quality and quantity of data available to estimate the parameter as opposed to the structure and parameterization of the model.

@@

\begin{figure}[!tbhp]
	\centering{\includegraphics[width=.6\textwidth,height=.6\textheight,keepaspectratio]{C:/Users/shyam/Documents/Courses/CHE1125Project/Results/ident/figure3/ident_analysis}}
	\caption{A flow diagram showing the methodology developed to establish practical identifiability of parameters in kinetic models of metabolism. a) The steps for the construction of a kinetic model of a metabolic network are shown. The choice of rate law formulations to describe metabolic fluxes influences the identification methodology. The identifiability of parameters for each flux can be establish independently. b) in silico data generation to test the identifiability of a given parameter is an important step that is also required for experimental design using the proposed methodology. c) The steps for identifiability analysis for a parameters of a single flux are shown. d) If any of the proposed in silico experiments results in non-identifiability of any parameter, then the number and nature of experiments to be performed to establish identifiability of the corresponding parameter can be determined by following the steps shown.}\label{fig:ident-flowchart}
\end{figure}

>>>>

\begin{document}
\paragraph{Outline:}
\begin{itemize}
	\item Importance of parameter values for estimating in vivo response using kinetic models of metabolism and consequently for metabolic network design using kinetic models of metabolism
	\item The need for parameter identifiability to determine unique and true parameter values from observed data
	\item Types and purpose of identifiability for parameters
	\item Methods for structural identifiability and existing methods for practical identifiability
	\item Lack of methods for practical identifiability and consequently experimental design (not covered?)
	\item work done in this paper for practical identifiability			
	\item scalability of computer algebra-based methods for structural identifiability (using CRNT to reduce networks to make structural identifiability scalable) (move to discussion - may be in discussion)
\end{itemize}
\section{Introduction:}
Kinetic models of metabolism can be used to study the dynamic characteristics of metabolic networks. In these models, ordinary differential equations (ode) are used to express the rate of change of metabolite concentrations ($x$) as a function of the reaction fluxes ($v$) in the metabolic network (Equation \ref{eq:kinstoich}). The matrix $\mathbf{S}$ in Equation (\ref{eq:kinstoich}a) defines the stoichiometric relationship between the fluxes and the concentrations of the metabolic network.
\begin{subequations}\label{eq:kinstoich}
	\begin{align}
	\dot{x} = \mathbf{S}v\\
	v = g(x, \theta, u)
	\end{align}
\end{subequations}
The expression for the nonlinear function ($g$) used to describe each reaction flux $v_i$ in a kinetic model (Equation \ref{eq:kinstoich}b) is dependent on the enzyme kinetic mechanism that is used to model the reaction \parencite{Heijnen2005, Link2014, Machado2011, Srinivasan2015}. Accordingly, $g$ is a nonlinear function of the metabolite concentrations, enzyme kinetic parameters ($\theta$) and other input concentrations ($u$).

The ability to predict the steady state and dynamic responses of metabolic networks, under in vivo conditions, to different perturbations is dependent on the numerical values of the enzyme kinetic parameter values ($\theta$) in Equation (\ref{eq:kinstoich}). Analyzing the ability of a metabolic network to exhibit dynamic characteristics like multiple steady states and oscillations, irrespective of the structure of the network, is one example where parameter values might play a crucial role \parencite{Srinivasan2015, Vital-Lopez2006}(Srinivasan et al., 2017?). The use of in vitro, or unreliable in vivo parameter estimates, reduces confidence in the model predicted behaviour. Consequently, the reduction in confidence hampers the use of these models to gain insight into the functioning of metabolic networks \parencite{Tran2008, Chakrabarti2013a}. The insights gained from the use of kinetic models are subsequently used to design changes to these metabolic networks to achieve various goals. These goals could either be to increase metabolite production for biosynthesis of different chemicals \parencite{Almquist2014, Khodayari2016, Costa2016, Andreozzi2016}(Srinivasan et al., 2017) or to find therapeutic targets to cure ailments \parencite{Apaolaza2017}. Hence, an increase in uncertainty in model predicted responses is also an obstacle for using the predicted responses as a basis for designing the metabolic networks to achieve these goals.

If all intracellular metabolite concentrations can be measured over a time course, a nonlinear programming problem can be formulated to estimate the enzyme kinetic parameters ($\theta$) in Equation (\ref{eq:kinstoich}), based on the measured data. The minimization of least square error between the measured ($x^*$) and modeled ($x$) concentrations can be used as an objective function (Equation \ref{eq:lsqopt}a) for the optimization problem (Equation \ref{eq:lsqopt}), with fixed upper ($\theta_u$) and lower ($\theta_l$) bounds for the parameter values (Equation \ref{eq:lsqopt}b). The difference between the data and the model estimate for each output and at each time point in the objective function is weighted by the variance in the experimental data $\sigma_{kl}^*$ for that corresponding variable and time point.
\begin{subequations}\label{eq:lsqopt}
	\begin{align}
	\underset{\theta}{\mathrm{min}} &\sum_{k=1}^{m}\sum_{l=1}^{d}\left(\frac{x_{kl}^*-x_{kl}}{\sigma_{kl}^*}\right)^2\\
	&\theta_l \le \theta \le \theta_u
	\end{align}
\end{subequations}

However, not all metabolite concentrations used in the model (Equation \ref{eq:kinstoich}) can be measured. Additionally, measurable fluxes in the metabolic network also need to be included as part of the parameter estimation problem. In such scenarios, the parameter estimation problem is modified to suit a new system of equations shown below (Equation \ref{eq:output}). The new system of equations is obtained by augmenting the original system (Equation \ref{eq:kinstoich}) with Equation (\ref{eq:output}c) that models the relationship between the measurable metabolite concentrations and fluxes ($y$) and the unmeasured concentrations ($x$) that are used in the original model (Equation \ref{eq:kinstoich}) above. Equation (\ref{eq:output}) now has additional parameters $\mu$ that define this relationship and also need to be estimated.
\begin{subequations}\label{eq:output}
	\begin{align}
	\dot{x} = \mathbf{S}v\\
	v = g(x, y, \theta, u)\\
	\dot{y} = h(x, y, \mu, u)
	\end{align}
\end{subequations}
In systems identification, the measured concentrations and fluxes ($y$) are called output or observed variables, and the unmeasured concentrations ($x$) are called the state variables. For estimating both $\theta$ and $\mu$, the metabolite concentrations $x$ in the optimization problem (Equation \ref{eq:lsqopt}) are substituted with the output variables $y$.

However, the ability to determine unique solutions to parameters $\theta$ and $\mu$ is governed by the identifiability of these parameters in the model \parencite{McLean2012}. The identifiability of parameters in nonlinear models can be classified into two categories: structural (or a priori) and practical (or posterior) identifiability. 	
Any system (Equation \ref{eq:output}) is said to be structurally identifiable if, for an input-output mapping defined by $y = \Phi(\mu,u)$ for at least one input function $u$, any two values of parameters $\mathbf{\mu}_1$ and $\mathbf{\mu}_2$ satisfy the relationship in Equation (\ref{eq:stident}) below.
\begin{align}\label{eq:stident}
\Phi(\mathbf{\mu}_1,u) = \Phi(\mathbf{\mu}_2,u) \iff \mathbf{\mu}_1 = \mathbf{\mu}_2
\end{align}
Accordingly, any system that has an infinite number of solutions to the parameter estimation problem for all input functions is said to structurally non-identifiable. So, the structural identifiability of parameters in a dynamic model helps establish the presence or absence of a relationship between the unobservable state variables and the observable output variables. Consequently, the effect of model structure and parameterization on the ability to infer true parameter values from experimental data is determined by the structural identifiability of the parameter.

Experimental data from many physical systems is usually noisy, and when parameters are estimated on the basis of noisy data, the ability to estimate unique parameter values to satisfy Equation (\ref{eq:stident}) is referred to as practical identifiability. So, the effect of the available experimental data on the ability to estimate unique parameter values is determined by the practical identifiability of the parameter. Accordingly, practical identifiability of a parameter is contingent upon the nature, quality and quantity of data available to estimate the parameter as opposed to the structure and parameterization of the model.

@@

\begin{figure}[!tbhp]
	\centering{\includegraphics[width=.6\textwidth,height=.6\textheight,keepaspectratio]{C:/Users/shyam/Documents/Courses/CHE1125Project/Results/ident/figure3/ident_analysis}}
	\caption{A flow diagram showing the methodology developed to establish practical identifiability of parameters in kinetic models of metabolism. a) The steps for the construction of a kinetic model of a metabolic network are shown. The choice of rate law formulations to describe metabolic fluxes influences the identification methodology. The identifiability of parameters for each flux can be established independently. b) The steps for identifiability analysis for parameters of a single flux are shown.}\label{fig:ident-flowchart}
\end{figure}
